RL-Musician: A Configurable Open-Source Tool for Species Counterpoint Composition

### 2020-04-11 docs/paper/paper.tex
docs: Describe Monte Carlo Beam Search in the draft

Currently, there are no perfect tools for algorithmic composition. Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that cannot be distinguished from works of a talented human composer. Thus, there is an open research problem. Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating, payne2019musenet} and reinforcement learning \cite{smith2012reinforcement} are applicable. Moreover, some researchers combine them \cite{jaques2016generating, kotecha2018bach, kumar2019polyphonic}. Meanwhile, fitness function optimization also yields prominent results\cite{herremans2012first, herremans2013fifth}.

@@

The closest to the current study works \cite{herremans2012first, herremans2013fifth} belong to the first group. A local search algorithm named Variable Neighborhood Search is used there in order to create first and fifth species counterpoint pieces that maximize fitness function. What these studies and the current one share in common is that fitness (reward) function is backed by the theory of species counterpoint. However, the current study is a bit closer to reinforcement learning, because composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions changes. Conversely, in \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations. As a result, pieces that break some fundamental rules of counterpoint are also evaluated there (they are marked as unfeasible), whereas here such pieces can not be created. This way of that, the core drawback of the discussed papers is that download links to software implementation named Optimuse are broken and so the results can not be reproduced.

There are other optimization-based approaches as well. Instead of local search metaheuristics, population-based metaheuristics (genetic algorithms, evolutionary strategies, cross-entropy method) or construction metaheuristics (ant colony algorithms) can be used. However, such papers are less relevant to the current study and so let us switch our attention to reinforcement learning. It can be applied to algorithmic composition either as a standalone paradigm or in connection with supervised deep learning. As for the first case, probably, there are no influential works and only separate studies like \cite{smith2012reinforcement, yi2007automatic} are available. Conversely, the second group recently gained attention and several interesting approaches were developed there.

@@

Consider a tree search problem.

@@

A software implementation of the above methodology in Python programming language is available on GitHub\footnote{\url{https://github.com/Nikolay-Lysenko/rl-musician}}. The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.2.0/}}. All important settings are placed to a separate configuration file and so it is easy to experiment with them.

>>>>

Currently, there are no perfect tools for algorithmic composition. Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that cannot be distinguished from works of a talented human composer. Thus, there is an open research problem. Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating, payne2019musenet} and reinforcement learning \cite{yi2007automatic, smith2012reinforcement} are applicable. Moreover, some researchers combine them \cite{jaques2016generating, kotecha2018bach, kumar2019polyphonic}. Meanwhile, fitness function optimization also yields prominent results\cite{herremans2012first, herremans2013fifth}.

@@

The closest to the current study works \cite{herremans2012first, herremans2013fifth} belong to the first group. A local search algorithm named Variable Neighborhood Search is used there in order to create first and fifth species counterpoint pieces that maximize fitness function. What these studies and the current one share in common is that fitness (reward) function is backed by the theory of species counterpoint. However, the current study is a bit closer to reinforcement learning, because composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions changes. Conversely, in \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations. As a result, pieces that break some fundamental rules of counterpoint are also evaluated there (they are marked as unfeasible), whereas here such pieces can not be created. This way of that, the core drawback of the two discussed papers is that download links to the software implementation named Optimuse are broken and so the results can not be reproduced.

There are other optimization-based approaches as well. Instead of local search metaheuristics, population-based metaheuristics (genetic algorithms, evolutionary strategies, cross-entropy method) or construction metaheuristics (ant colony algorithms) can be used. However, such papers are less relevant to the current study and so let us switch our attention to reinforcement learning. It can be applied to algorithmic composition either as a standalone paradigm or in connection with supervised deep learning. As for the first case, probably, there are no influential works and only separate studies like \cite{yi2007automatic, smith2012reinforcement} are available. Conversely, the second group recently gained attention and several interesting approaches were developed there.

@@

Consider a tree search problem. To be more detailed, suppose that there is a tree such that every its leaf is mapped to a reward and the goal is to find the leaf with the maximum reward. However, leaves and corresponding to them rewards are not known initially. Information about a leaf becomes available only after a path from the root of the tree to this leaf is constructed. %In other words, the goal is to find optimum sequence of nodes starting at the root and ending at a leaf.

The described in Section \ref{sec:introduction} procedure of first species counterpoint writing can be considered a tree search problem. The root is the initial state of an environment. Each valid option to fill the current measure is an edge from the current node of the tree to a new node. Leaves are pieces where all measures are filled.

If size of a tree is small enough, exhaustive search is possible. There are well-known classical algorithms such as depth-first search (DFS) that allow iterating over all paths from the root to leaves. Unfortunately, trees are large in numerous real problems. For instance, a rough estimate of total number of leaves for first species counterpoint with 2 voices and 16 measures is about $7^{14}$ which is equal to several hundreds of billions.

An obvious idea for a large tree is to use random search. To create a path, every next node is chosen at random until a leaf is reached. After specified number of paths is created, the best one amongst them is chosen as an approximate solution.

This idea can be further improved in the following way. Let an optimum path be constructed sequentially with adding one new node per a series of random trials. If so, there is a stub initially containing only the root. At every iteration, there are several trials to continue stub at random until a leaf is reached. The best one amongst them (and amongst trials from previous iterations if applicable) is chosen and the first non-stub node from there is added to the stub. This method can be called Monte Carlo Search. The gain in comparison with plain random search is due to allocating more exploration to subtrees that contain the current best leaf.

Above version of Monte Carlo Search is greedy, because it ignores that some subtrees might be underexplored. A well-known method that explores more than greedy search, but uses less computational resources than exhaustive search, is called beam search. Instead of just one stub, $w$ distinct stubs are constructed and used, where $w$ is a hyperparameter named beam width. It is natural to call this method Monte Carlo Beam Search.

Probably, the term Monte Carlo Beam Search was coined in \cite{cazenave2012beam}. However, it has different meaning there. In \cite{cazenave2009nested}, the same author defined Nested Monte Carlo Search as a combination of Monte Carlo Search with DFS. Given a current stub, all its extensions by $n$ new nodes (where $n$ is a hyperparameter named level of nestedness) are found with DFS and then continued at random. So pre-defined number of trials per iteration is absent, because number of trials is equal to number of extensions by $n$ nodes ahead. Comparing with Monte Carlo Search, Nested Monte Carlo Search distributes exploration more evenly between short extensions of a stub. This way or that, in \cite{cazenave2012beam}, Monte Carlo Beam Search is defined as beam search based on Nested Monte Carlo Search.

Nevertheless, here, beam search is based on Monte Carlo Search. The reason for it is quite mundane -- parallel implementation of DFS looks like overcomplication that may yield too few. The second difference with \cite{cazenave2012beam} is that dependence of trials number on stub length is introduced. The longer the stub is, the smaller a subtree to be explored is. Therefore, it is better to reallocate computational resources in order to explore more when a stub is short and to explore less when a stub is long.

@@

A software implementation of the above methodology in Python programming language is available on GitHub\footnote{\url{https://github.com/Nikolay-Lysenko/rl-musician}}. The code has built-in documentation, is covered with unit tests, and is released as a package on PyPI\footnote{\url{https://pypi.org/project/rl-musician/0.3.0/}}. All important settings are placed to a separate configuration file and so it is easy to experiment with them.
